Endogenous Macrodynamics in Algorithmic Recourse

### 2022-08-16 paper/paper.tex
progress on results, formatting and comments

% *** CITATION PACKAGES ***
%
%\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of the IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off
% such as if a citation ever needs to be enclosed in parenthesis.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 5.0 (2009-03-20) and later if using hyperref.sty.
% The latest version can be obtained at:
% http://www.ctan.org/pkg/cite
% The documentation is contained in the cite.sty file itself.

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely been limited to the static setting and focused on single individuals: given some estimated model the goal is to find valid counterfactuals for individual instance that fulfill various desiderata. The ability of such counterfactuals to handle dynamics like data and model drift remains a largely unexplored research challenge at this point. There has also been surprisingly little work on the related question of how the actual implementation of recourse by one individual may affect other individuals. Through this work we aim to close that gap by systematizing and extending existing knowledge. We first show that many of the existing methodologies can be collectively described by a generalized framework. We then argue that the existing framework fails to account for a hidden external cost of recourse, that only reveals itself when studying the endogenous dynamics of recourse at the group level. Through simulation experiments involving various popular counterfactual generators and several benchmark datasets, we generate a total XX million Counterfactual Explanations and study the resulting domain and model shifts. We find that the induced shifts are substantial enough to likely impede the applicability Algorithmic Recourse in practice. Fortunately, we find various potential mitigation strategies that can be used in combination with existing approaches. Our simulation framework for studying recourse dynamics is fast and open-sourced.
\end{abstract}

@@

Recent advances in Artificial Intelligence (AI) have propelled its adoption in scientific domains outside of Computer Science including Healthcare, Bioinformatics, Genetics and the Social Sciences. While this has in many cases brought benefits in terms of efficiency, state-of-the-art models like Deep Neural Networks (DNN) have also given rise a new type of problem in the context of data-driven decision-making. They are essentially \textbf{black boxes}: so complex, opaque and underspecified in the data that it is often impossible to understand how they actually arrive at their decision without auxiliary tools. Despite this shortcoming, black-box models have grown in popularity in recent years and have at times created undesirable societal outcomes (O'neil 2016). The scientific community has tackled this issue from two different angles: while some have appealed for a strict focus on inherently interpretable models (Rudin 2019), others have investigated different ways to explain the behaviour of black-box models. These two sub-domains can be broadly referred to as \textbf{interpretable AI} and \textbf{explainable AI} (XAI), respectively.

@@

Existing work in this field has largely worked in a static setting: various approaches have been proposed to generate counterfactuals for a given individual that is subject to some pre-trained model. More recent work has compared different approaches within this static setting (Pawelczyk et al. 2021). In this work, we go one step further and ask ourselves: what happens if recourse is provided and implemented repeatedly? What types of dynamics are introduced and how do different counterfactual generators compare in this context?

@@

Figure \ref{fig:poc} illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by Wachter, Mittelstadt, and Russell (2017): the implementation of AR for a subset of individuals leads to a domain shift (b), which in turn triggers a model shift (c). As this game of implementing AR and updating the classifier is repeated, the decision boundary moves away from training samples that were originally in the target class (d). We refer to these types of dynamics as \textbf{endogenous} because they are induced by the implementation of recourse itself. The term \textbf{macrodynamics} is borrowed from the economics literature and used to describe processes involving whole groups or societies.

@@

To the best of our knowledge this is the first work investigating endogenous macrodynamics in AR. Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on Algorithmic Recourse extending our focus from single individuals to groups. Secondly, we introduce an experimental framework extending previous work by Altmeyer (2022), which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized. Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including Wachter, Mittelstadt, and Russell (2017), Schut et al. (2021), Joshi et al. (2019), Mothilal, Sharma, and Tan (2020) and Antorán et al. (2020). To this end we propose a number of novel evaluation metrics that can be used to quantify and benchmark the macrodynamics introduced by the different generators. Finally, we also discuss what drives endogenous dynamics and propose strategies to mitigate them.

@@

A framework for Counterfactual Explanations was first proposed in 2017 by Wachter, Mittelstadt, and Russell (2017) and has served as the baseline for most methodologies that have been proposed since then. Let \(M: \mathcal{X} \mapsto \mathcal{Y}\) denote some pre-trained model that maps from inputs \(X \in \mathcal{X}\) to outputs \(Y \in \mathcal{Y}\). Then we are interested in minimizing the cost\footnote{Equivalently, others have referred to this quantity as \emph{complexity} or simply \emph{distance}} \(H=h(x\prime)\) incurred by individual \(x\) when moving to a counterfactual state \(x\prime\) such that the predicted outcome \(M(x\prime)\) corresponds to some target outcome \(y^*\):

\begin{equation}
\min_{x\prime \in \mathcal{x}} h(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = y^* \label{eq:obj}
\end{equation}

@@

\begin{equation}
x\prime = \arg \min_{x\prime}  \ell(M(x\prime),y^*) + \lambda h(x\prime) \label{eq:solution}
\end{equation}

In the baseline work (Wachter, Mittelstadt, and Russell 2017), the cost function is proxied by some distance metric based on the simple intuition that perturbations of \(x\) are costly to the individual. For models that are differentiable and produce smooth predictions, \eqref{eq:solution} can be solved through gradient descent. This summarizes the approach followed in Wachter, Mittelstadt, and Russell (2017) which we shall refer to simply as \textbf{Wachter} - the name of the first author - in the remainder of this paper.

Many approaches for the generation of Algorithmic Recourse have been described in the literature since 2017. An October 2020 survey by Karimi et al.~layed out 60 algorithms that have been proposed since 2014 (Karimi et al. 2020). Another survey published around the same time by Verma et al.~described 29 algorithms (Verma, Dickerson, and Hines 2020). Different approaches vary primarily in terms of the objective functions they impose, how they optimize said objective (from brute force through gradient-based approaches to graph traversal algorithms), and how the ensure that certain requirements for CE are met. Regarding the latter, the literature has produced an extensive list of desiderata each addressing different needs. To name but a few, we are interested in generating counterfactuals that close (Wachter, Mittelstadt, and Russell 2017), actionable (Ustun, Spangher, and Liu 2019), realistic Schut et al. (2021), sparse, diverse (Mothilal, Sharma, and Tan 2020) and if possible causally founded (Karimi, Schölkopf, and Valera 2021).

Efforts so far have largely been directed at improving the quality of Counterfactual Explanations within a static context: given some pre-trained classifier \(M: \mathcal{X} \mapsto \mathcal{Y}\), we are interested in generating one or multiple meaningful Counterfactual Explanations for some individual characterized by \(x\). The ability of Counterfactual Explanations to handle dynamics like data and model shifts remains a largely unexplored research challenge at this point (Verma, Dickerson, and Hines 2020). We have been able to identify only one recent work that considers the implications of \textbf{exogenous} domain and model shifts in the context of AR (Upadhyay, Joshi, and Lakkaraju 2021). Exogenous shifts are strictly of external origin. For example, they might stem from data correction, temporal shifts or geospatial changes (Upadhyay, Joshi, and Lakkaraju 2021). The authors of (Upadhyay, Joshi, and Lakkaraju 2021) propose ROAR - a framework for Algorithmic Recourse that evidently improves robustness to such exogenous shifts.

As mentioned earlier, research has so far also generally focused on generating counterfactuals for single individuals or instances. We have been able to identify only one existing work that investigates black-box model behavior towards a group of individuals (Carrizosa, Ramırez-Ayerbe, and Romero 2021). The authors propose an optimization framework that generates collective counterfactuals. We provide a motivation for doing so from the perspective of endogenous macrodynamics of Algorithmic Recourse.

@@

Much attention has been paid to the detection of dataset shifts -- situations where the distribution of data changes over time. Rabanser et al.~suggest a framework to detect data drift from a minimal number of samples through the application of two-sample tests (Rabanser, Günnemann, and Lipton 2019). This task is a generalization of the anomaly detection problem for large datasets, which aims to answer the question if two sets of samples could have been generated from the same probability distribution. Numerous approaches to anomaly detection have been summarized (Chandola, Banerjee, and Kumar 2009). Another well-established research topic is that of concept drift -- situations where external variables influence the patterns between the input and the output of a model (Widmer and Kubat 1996). For instance, Gama et al.~offer a review of the adaptive learning techniques which can handle concept drift (Gama et al. 2014). Less previous work is available on the related topic of model drift - changes in model performance over time. Nelson et al.~review how resistant different machine learning models are to the model drift (Nelson et al. 2015). Ackerman et al.~offer a method to detect changes in model performance when ground truth is not available (Ackerman et al. 2021).

In the context of Algorithmic Recourse, domain and model shifts were first brought up by the authors behind ROAR (Upadhyay, Joshi, and Lakkaraju 2021). In their work they refer to model shifts as simply any perturbation \(\Delta\) to the parameters of the model in question: \(M\). While this also sets the baseline for our analysis here, it is worth noting that in Upadhyay, Joshi, and Lakkaraju (2021) these perturbations are mechanically introduced. In contrast we are interested in quantifying model shifts that arise endogenously as part of a dynamic recourse process. In addition to quantifying the magnitude of shifts \(\Delta\), we aim to also analyse the characteristics of changes to the model, such as the position of the decision boundary and the overall decisiveness of the model. We have not been able to identify previous work on this topic.

@@

Despite the large and growing number of approaches to counterfactual search, there have been surprisingly few benchmark studies that compare different methodologies. This may be partially due to limited software availability in this space. Recent work has started to address this gap: firstly, Oliveira and Martens (2021) run a large benchmarking study using different algorithmic aproaches and numerous tabular datasets; secondly, Pawelczyk et al. (2021) introduce a Python framework - CARLA - that can be used to apply and benchmark different methodologies; finally, \texttt{CounterfactualExplanations.jl} (Altmeyer 2022) provides an extensible, fast and language-agnostic implementation in Julia. Since the experiments presented here involve extensive simulations, we have relied on and extended the Julia implementation due to the associated performance benefits. In particular, we have built a framework on top of \texttt{CounterfactualExplanations.jl} that extends the functionality from static benchmarks to simulation experiments: \href{(https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md)}{\texttt{AlgorithmicRecourseDynamics.jl}}\footnote{The code is available from the following anonymized Github repository: \url{https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md}.}. The core concepts implemented in that package reflect what is presented in Section \ref{method-2} of this paper.

@@

We have chosen to focus on gradient-based counterfactual search for two reasons: firstly, they can be seen as direct descendants of our baseline method - Wachter; secondly, gradient-based search is particularly well-suited for differentiable black-box models like deep neural networks, which we focus on in this work. In particular, we include include the following generators in our simulation experiments below: \textbf{REVISE} (Joshi et al. 2019), \textbf{CLUE} (Antorán et al. 2020), \textbf{DiCE} (Mothilal, Sharma, and Tan 2020) and a greedy approach that relies on probabilistic models (Schut et al. 2021). Our motivation for including these different generators in our analysis, is that they all offer slightly different approaches to generate meaningful counterfactuals for differentiable black-box models. We hypothesize that generating more \textbf{meaningful} counterfactuals should mitigate the endogenous dynamics illustrated in Figure \ref{fig:poc} in Section \ref{intro}. This intuition stems from the underlying idea that more meaningful counterfactuals are generated by the same or at least a very similar data generating process as the training data. All else equal, counterfactuals that fulfill this basic requirement should be less prone to trigger domain and model shifts.

@@

\begin{equation}
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)}+ \lambda {h(f(s_k^\prime)) }  \right\} \label{eq:general}
\end{aligned} 
\end{equation}

Here \(\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K\) is the stacked \(K\)-dimensional array of counterfactual states and \(f: \mathcal{S} \mapsto \mathcal{X}\) maps from the counterfactual state space to the feature space. In Wachter, the state space is the feature space: \(f\) is just the identity function and the number of counterfactuals \(K\) is equal to one. Both REVISE and CLUE search counterfactuals in some latent embedding \(S \subset \mathcal{S}\) instead of the feature space directly. The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of \(X\). In this case \(f\) in Equation \eqref{eq:general} corresponds to the decoder part of the generative model, in other words the function that maps back from the latent embedding to the feature space. Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP (Joshi et al. 2019).

CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question, \(M\). To quantify predictive uncertainty the authors rely on entropy estimates for probabilistic models. The approach proposed by Schut et al. (2021), which we shall refer to as \textbf{Greedy}, also works with the subclass of models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that can produce predictive uncertainty estimates. The authors show that in this setting the cost function \(h(\cdot)\) in Equation \eqref{eq:general} is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA). Schut et al. (Schut et al. 2021) also show that by maximizing the predicted probability of \(x^\prime\) being assigned to target class \(y^*\) we also implicitly minimize predictive entropy - as in CLUE. In that sense, CLUE can be seen as equivalent to REVISE in the Bayesian context and we shall therefore refer to both approaches collectively as \textbf{Latent Space} generators\footnote{In fact, there are a number of other recently proposed approaches to counterfactual search that also broadly fall in this same category. They largely differ with respect to the chosen generative model: for example, the generator proposed by Dombrowski, Gerken, and Kessel (2021) relies on normalizing flows.}.

Finally, DiCE (Mothilal, Sharma, and Tan 2020) distinguishes itself from all other generators considered here in that it aims to generate a diverse set of \(K>1\) counterfactuals. Wachter et al. (Wachter, Mittelstadt, and Russell 2017) show that diverse outcomes can in principal be achieved simply rerunning counterfactual search multiple times using stochastic gradient descent (or by randomly initializing the counterfactual). In Mothilal, Sharma, and Tan (2020) diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function \(h(\mathbf{s}^\prime)\) and therebt produce counterfactuals \(s_1, ... , s_K\) that look as different from each other as possible. The implementation of DiCE in the our library of choice - \texttt{CounterfactualExplanations.jl} - uses that exact approach. It is worth noting that for \(k=1\), DiCE reduces to Wachter since the DDP is constant and therefore does not affect the objective function in Equation \eqref{eq:general}.

@@

\begin{equation}
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\ell(M(f(s_k^\prime)),y^*)} \\ &+ \lambda_1 {h(f(s_k^\prime)) } + \lambda_2 {g(f(s_k^\prime))}  \label{eq:collective}
\end{aligned} 
\end{equation}

Here \(h(f(s_k^\prime))\) denotes the proxy for private costs faced by the individual as before and \(\lambda_1\) governs to what extent that private cost ought to be penalized. The newly introduced term \(g(f(s_k^\prime))\) is meant to capture and address external costs incurred by the collective of individuals in response to changes in \(\mathbf{s}^\prime\). The underlying concept of private and external costs is borrowed from Economics and well-established in that field: when the decisions or actions by some individual market participant generate external costs, then the market is said to suffer from negative externalities and considered inefficient (Pindyck and Rubinfeld 2014). We think that this concept describes the endogenous dynamics of algorithmic recourse oberserved here very well. As with individual recourse, the exact choice of \(g(\cdot)\) is not obvious, nor do we intend to provide a definite answer in this work, if such even exists. That being said, we do propose a few potential mitigation strategies in Section \ref{mitigate}.

@@

\{\{\textless{} include sections/appendix.qmd \textgreater\}\}

>>>>

% *** CITATION PACKAGES ***
%
\usepackage{cite}
%\usepackage{cite}
% cite.sty was written by Donald Arseneau
% V1.6 and later of IEEEtran pre-defines the format of the cite.sty package
% \cite{} output to follow that of the IEEE. Loading the cite package will
% result in citation numbers being automatically sorted and properly
% "compressed/ranged". e.g., [1], [9], [2], [7], [5], [6] without using
% cite.sty will become [1], [2], [5]--[7], [9] using cite.sty. cite.sty's
% \cite will automatically add leading space, if needed. Use cite.sty's
% noadjust option (cite.sty V3.8 and later) if you want to turn this off
% such as if a citation ever needs to be enclosed in parenthesis.
% cite.sty is already installed on most LaTeX systems. Be sure and use
% version 5.0 (2009-03-20) and later if using hyperref.sty.
% The latest version can be obtained at:
% http://www.ctan.org/pkg/cite
% The documentation is contained in the cite.sty file itself.

@@

% As a general rule, do not put math, special symbols or citations
% in the abstract
\begin{abstract}
Existing work on Counterfactual Explanations (CE) and Algorithmic Recourse (AR) has largely been limited to the static setting and focused on single individuals: given some estimated model the goal is to find valid counterfactuals for individual instance that fulfill various desiderata. The ability of such counterfactuals to handle dynamics like data and model drift remains a largely unexplored research challenge at this point. There has also been surprisingly little work on the related question of how the actual implementation of recourse by one individual may affect other individuals. Through this work we aim to close that gap by systematizing and extending existing knowledge. We first show that many of the existing methodologies can be collectively described by a generalized framework. We then argue that the existing framework fails to account for a hidden external cost of recourse, that only reveals itself when studying the endogenous dynamics of recourse at the group level. Through simulation experiments involving various state-of-the-art counterfactual generators and several benchmark datasets, we generate a total of XX million counterfactuals and study the resulting domain and model shifts. We find that the induced shifts are substantial enough to likely impede the applicability Algorithmic Recourse in situations that involve competition for scarce resources. Fortunately, we find various potential mitigation strategies that can be used in combination with existing approaches. Our simulation framework for studying recourse dynamics is fast and open-sourced.
\end{abstract}

@@

Recent advances in Artificial Intelligence (AI) have propelled its adoption in scientific domains outside of Computer Science including Healthcare, Bioinformatics, Genetics and the Social Sciences. While this has in many cases brought benefits in terms of efficiency, state-of-the-art models like Deep Neural Networks (DNN) have also given rise a new type of problem in the context of data-driven decision-making. They are essentially \textbf{black boxes}: so complex, opaque and underspecified in the data that it is often impossible to understand how they actually arrive at their decision without auxiliary tools. Despite this shortcoming, black-box models have grown in popularity in recent years and have at times created undesirable societal outcomes \protect\hyperlink{ref-o2016weapons}{{[}1{]}}. The scientific community has tackled this issue from two different angles: while some have appealed for a strict focus on inherently interpretable models \protect\hyperlink{ref-rudin2019stop}{{[}2{]}}, others have investigated different ways to explain the behaviour of black-box models. These two sub-domains can be broadly referred to as \textbf{interpretable AI} and \textbf{explainable AI} (XAI), respectively.

@@

Existing work in this field has largely worked in a static setting: various approaches have been proposed to generate counterfactuals for a given individual that is subject to some pre-trained model. More recent work has compared different approaches within this static setting \protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}}. In this work, we go one step further and ask ourselves: what happens if recourse is provided and implemented repeatedly? What types of dynamics are introduced and how do different counterfactual generators compare in this context?

@@

Figure \ref{fig:poc} illustrates this idea for a binary problem involving a probabilistic classifier and the counterfactual generator proposed by \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}: the implementation of AR for a subset of individuals leads to a domain shift (b), which in turn triggers a model shift (c). As this game of implementing AR and updating the classifier is repeated, the decision boundary moves away from training samples that were originally in the target class (d). We refer to these types of dynamics as \textbf{endogenous} because they are induced by the implementation of recourse itself. The term \textbf{macrodynamics} is borrowed from the economics literature and used to describe processes involving whole groups or societies.

@@

To the best of our knowledge this is the first work investigating endogenous macrodynamics in AR. Our contributions to the state of knowledge are as follows: firstly, we posit a compelling argument that calls for a novel perspective on Algorithmic Recourse extending our focus from single individuals to groups. Secondly, we introduce an experimental framework extending previous work by \protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}5{]}}, which enables us to study macrodynamics of Algorithmic Recourse through simulations that can be fully parallelized. Thirdly, we use this framework to provide a first in-depth analysis of endogenous recourse dynamics induced by various popular counterfactual generators including \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, \protect\hyperlink{ref-joshi2019towards}{{[}7{]}}, \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and \protect\hyperlink{ref-antoran2020getting}{{[}9{]}}. To this end we propose a number of novel evaluation metrics that can be used to quantify and benchmark the macrodynamics introduced by the different generators. Finally, we also discuss what drives endogenous dynamics and propose strategies to mitigate them.

@@

A framework for Counterfactual Explanations was first proposed in 2017 by \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} and has served as the baseline for most methodologies that have been proposed since then. Let \(M: \mathcal{X} \mapsto \mathcal{Y}\) denote some pre-trained model that maps from inputs \(X \in \mathcal{X}\) to outputs \(Y \in \mathcal{Y}\). Then we are interested in minimizing the cost\footnote{Equivalently, others have referred to this quantity as \emph{complexity} or simply \emph{distance}} \(C=\text{cost}(x\prime)\) incurred by individual \(x\) when moving to a counterfactual state \(x\prime\) such that the predicted outcome \(M(x\prime)\) corresponds to some target outcome \(y^*\):

\begin{equation}
\min_{x\prime \in \mathcal{X}} \text{cost}(x\prime) \ \ \ \mbox{s. t.} \ \ \ M(x\prime) = y^* \label{eq:obj}
\end{equation}

@@

\begin{equation}
x\prime = \arg \min_{x\prime}  \text{yloss}(M(x\prime),y^*) + \lambda \text{cost}(x\prime) \label{eq:solution}
\end{equation}

In the baseline work \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, the cost function is proxied by some distance metric based on the simple intuition that perturbations of \(x\) are costly to the individual. For models that are differentiable and produce smooth predictions, \eqref{eq:solution} can be solved through gradient descent. This summarizes the approach followed in \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} which we shall refer to simply as \textbf{Wachter} - the name of the first author - in the remainder of this paper.

Many approaches for the generation of Algorithmic Recourse have been described in the literature since 2017. An October 2020 survey by Karimi et al.~layed out 60 algorithms that have been proposed since 2014 \protect\hyperlink{ref-karimi2020survey}{{[}10{]}}. Another survey published around the same time by Verma et al.~described 29 algorithms \protect\hyperlink{ref-verma2020counterfactual}{{[}11{]}}. Different approaches vary primarily in terms of the objective functions they impose, how they optimize said objective (from brute force through gradient-based approaches to graph traversal algorithms), and how the ensure that certain requirements for CE are met. Regarding the latter, the literature has produced an extensive list of desiderata each addressing different needs. To name but a few, we are interested in generating counterfactuals that close \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}}, actionable \protect\hyperlink{ref-ustun2019actionable}{{[}12{]}}, realistic \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, sparse, diverse \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and if possible causally founded \protect\hyperlink{ref-karimi2021algorithmic}{{[}13{]}}.

Efforts so far have largely been directed at improving the quality of Counterfactual Explanations within a static context: given some pre-trained classifier \(M: \mathcal{X} \mapsto \mathcal{Y}\), we are interested in generating one or multiple meaningful Counterfactual Explanations for some individual characterized by \(x\). The ability of Counterfactual Explanations to handle dynamics like data and model shifts remains a largely unexplored research challenge at this point \protect\hyperlink{ref-verma2020counterfactual}{{[}11{]}}. We have been able to identify only one recent work that considers the implications of \textbf{exogenous} domain and model shifts in the context of AR \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. Exogenous shifts are strictly of external origin. For example, they might stem from data correction, temporal shifts or geospatial changes \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. The authors of \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}} propose ROAR - a framework for Algorithmic Recourse that evidently improves robustness to such exogenous shifts.

As mentioned earlier, research has so far also generally focused on generating counterfactuals for single individuals or instances. We have been able to identify only one existing work that investigates black-box model behavior towards a group of individuals \protect\hyperlink{ref-carrizosa2021generating}{{[}15{]}}. The authors propose an optimization framework that generates collective counterfactuals. We provide a motivation for doing so from the perspective of endogenous macrodynamics of Algorithmic Recourse.

@@

Much attention has been paid to the detection of dataset shifts -- situations where the distribution of data changes over time. Rabanser et al.~suggest a framework to detect data drift from a minimal number of samples through the application of two-sample tests \protect\hyperlink{ref-rabanser2019failing}{{[}16{]}}. This task is a generalization of the anomaly detection problem for large datasets, which aims to answer the question if two sets of samples could have been generated from the same probability distribution. Numerous approaches to anomaly detection have been summarized \protect\hyperlink{ref-chandola2009anomaly}{{[}17{]}}. Another well-established research topic is that of concept drift -- situations where external variables influence the patterns between the input and the output of a model \protect\hyperlink{ref-widmer1996learning}{{[}18{]}}. For instance, Gama et al.~offer a review of the adaptive learning techniques which can handle concept drift \protect\hyperlink{ref-gama2014survey}{{[}19{]}}. Less previous work is available on the related topic of model drift - changes in model performance over time. Nelson et al.~review how resistant different machine learning models are to the model drift \protect\hyperlink{ref-nelson2015evaluating}{{[}20{]}}. Ackerman et al.~offer a method to detect changes in model performance when ground truth is not available \protect\hyperlink{ref-ackerman2021machine}{{[}21{]}}.

In the context of Algorithmic Recourse, domain and model shifts were first brought up by the authors behind ROAR \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}}. In their work they refer to model shifts as simply any perturbation \(\Delta\) to the parameters of the model in question: \(M\). While this also sets the baseline for our analysis here, it is worth noting that in \protect\hyperlink{ref-upadhyay2021towards}{{[}14{]}} these perturbations are mechanically introduced. In contrast we are interested in quantifying model shifts that arise endogenously as part of a dynamic recourse process. In addition to quantifying the magnitude of shifts \(\Delta\), we aim to also analyse the characteristics of changes to the model, such as the position of the decision boundary and the overall decisiveness of the model. We have not been able to identify previous work on this topic.

@@

Despite the large and growing number of approaches to counterfactual search, there have been surprisingly few benchmark studies that compare different methodologies. This may be partially due to limited software availability in this space. Recent work has started to address this gap: firstly, \protect\hyperlink{ref-de2021framework}{{[}22{]}} run a large benchmarking study using different algorithmic aproaches and numerous tabular datasets; secondly, \protect\hyperlink{ref-pawelczyk2021carla}{{[}3{]}} introduce a Python framework - CARLA - that can be used to apply and benchmark different methodologies; finally, \texttt{CounterfactualExplanations.jl} \protect\hyperlink{ref-altmeyer2022CounterfactualExplanations}{{[}5{]}} provides an extensible, fast and language-agnostic implementation in Julia. Since the experiments presented here involve extensive simulations, we have relied on and extended the Julia implementation due to the associated performance benefits. In particular, we have built a framework on top of \texttt{CounterfactualExplanations.jl} that extends the functionality from static benchmarks to simulation experiments: \href{(https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md)}{\texttt{AlgorithmicRecourseDynamics.jl}}\footnote{The code is available from the following anonymized Github repository: \url{https://anonymous.4open.science/r/AlgorithmicRecourseDynamics/README.md}.}. The core concepts implemented in that package reflect what is presented in Section \ref{method-2} of this paper.

@@

We have chosen to focus on gradient-based counterfactual search for two reasons: firstly, they can be seen as direct descendants of our baseline method - Wachter; secondly, gradient-based search is particularly well-suited for differentiable black-box models like deep neural networks, which we focus on in this work. In particular, we include include the following generators in our simulation experiments below: \textbf{REVISE} \protect\hyperlink{ref-joshi2019towards}{{[}7{]}}, \textbf{CLUE} \protect\hyperlink{ref-antoran2020getting}{{[}9{]}}, \textbf{DiCE} \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} and a greedy approach that relies on probabilistic models \protect\hyperlink{ref-schut2021generating}{{[}6{]}}. Our motivation for including these different generators in our analysis, is that they all offer slightly different approaches to generate meaningful counterfactuals for differentiable black-box models. We hypothesize that generating more \textbf{meaningful} counterfactuals should mitigate the endogenous dynamics illustrated in Figure \ref{fig:poc} in Section \ref{intro}. This intuition stems from the underlying idea that more meaningful counterfactuals are generated by the same or at least a very similar data generating process as the training data. All else equal, counterfactuals that fulfill this basic requirement should be less prone to trigger domain and model shifts.

@@

\begin{equation}
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}} \left\{ \sum_{k=1}^{K} {\text{yloss}(M(f(s_k^\prime)),y^*)}+ \lambda {\text{cost}(f(s_k^\prime)) }  \right\} \label{eq:general}
\end{aligned} 
\end{equation}

Here \(\mathbf{s}^\prime=\left\{s_k^\prime\right\}_K\) is the stacked \(K\)-dimensional array of counterfactual states and \(f: \mathcal{S} \mapsto \mathcal{X}\) maps from the counterfactual state space to the feature space. In Wachter, the state space is the feature space: \(f\) is just the identity function and the number of counterfactuals \(K\) is equal to one. Both REVISE and CLUE search counterfactuals in some latent embedding \(S \subset \mathcal{S}\) instead of the feature space directly. The latent embedding is learned by a separate generative model that is tasked with learning the data generating process (DGP) of \(X\). In this case \(f\) in Equation \eqref{eq:general} corresponds to the decoder part of the generative model, in other words the function that maps back from the latent embedding to the feature space. Provided the generative model is well-specified, traversing the latent embedding typically results in realistic and plausible counterfactuals, because they are implicitly generated by the (learned) DGP \protect\hyperlink{ref-joshi2019towards}{{[}7{]}}.

CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question, \(M\). To quantify predictive uncertainty the authors rely on entropy estimates for probabilistic models. The approach proposed by \protect\hyperlink{ref-schut2021generating}{{[}6{]}}, which we shall refer to as \textbf{Greedy}, also works with the subclass of models \(\tilde{\mathcal{M}}\subset\mathcal{M}\) that can produce predictive uncertainty estimates. The authors show that in this setting the cost function \(\text{cost}(\cdot)\) in Equation \eqref{eq:general} is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA). Schut et al. \protect\hyperlink{ref-schut2021generating}{{[}6{]}} also show that by maximizing the predicted probability of \(x^\prime\) being assigned to target class \(y^*\) we also implicitly minimize predictive entropy - as in CLUE. In that sense, CLUE can be seen as equivalent to REVISE in the Bayesian context and we shall therefore refer to both approaches collectively as \textbf{Latent Space} generators\footnote{In fact, there are a number of other recently proposed approaches to counterfactual search that also broadly fall in this same category. They largely differ with respect to the chosen generative model: for example, the generator proposed by \protect\hyperlink{ref-dombrowski2021diffeomorphic}{{[}23{]}} relies on normalizing flows.}.

Finally, DiCE \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} distinguishes itself from all other generators considered here in that it aims to generate a diverse set of \(K>1\) counterfactuals. Wachter et al. \protect\hyperlink{ref-wachter2017counterfactual}{{[}4{]}} show that diverse outcomes can in principal be achieved simply rerunning counterfactual search multiple times using stochastic gradient descent (or by randomly initializing the counterfactual). In \protect\hyperlink{ref-mothilal2020explaining}{{[}8{]}} diversity is explicitly proxied via Determinantal Point Processes (DDP): the authors simply introduce DDP as a component of the cost function \(\text{cost}(\mathbf{s}^\prime)\) and therebt produce counterfactuals \(s_1, ... , s_K\) that look as different from each other as possible. The implementation of DiCE in the our library of choice - \texttt{CounterfactualExplanations.jl} - uses that exact approach. It is worth noting that for \(k=1\), DiCE reduces to Wachter since the DDP is constant and therefore does not affect the objective function in Equation \eqref{eq:general}.

@@

\begin{equation}
\begin{aligned}
\mathbf{s}^\prime &= \arg \min_{\mathbf{s}^\prime \in \mathcal{S}}  \sum_{k=1}^{K} {\text{yloss}(M(f(s_k^\prime)),y^*)} \\ &+ \lambda_1 {\text{cost}(f(s_k^\prime))} + \lambda_2 {\text{extcost}(f(s_k^\prime))}  \label{eq:collective}
\end{aligned} 
\end{equation}

Here \(\text{cost}(f(s_k^\prime))\) denotes the proxy for private costs faced by the individual as before and \(\lambda_1\) governs to what extent that private cost ought to be penalized. The newly introduced term \(\text{extcost}(f(s_k^\prime))\) is meant to capture and address external costs incurred by the collective of individuals in response to changes in \(\mathbf{s}^\prime\). The underlying concept of private and external costs is borrowed from Economics and well-established in that field: when the decisions or actions by some individual market participant generate external costs, then the market is said to suffer from negative externalities and considered inefficient \protect\hyperlink{ref-pindyck2014microeconomics}{{[}24{]}}. We think that this concept describes the endogenous dynamics of algorithmic recourse oberserved here very well. As with individual recourse, the exact choice of \(\text{extcost}(\cdot)\) is not obvious, nor do we intend to provide a definite answer in this work, if such even exists. That being said, we do propose a few potential mitigation strategies in Section \ref{mitigate}.

@@
